---
title: "Do Double Stuf Oreos Have Twice the Creme Filling of Regular Oreos?"
author: "Student E"
date: "9/4/2026"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

# Research question

Do Double Stuf Oreos have **twice the mean creme filling mass** of regular Oreos?

The response is filling mass (grams), and the explanatory variable is Oreo type
(Regular or Double Stuf). The data are available from the STAT 461 repository at
<https://raw.githubusercontent.com/neilhatfield/STAT461/master/dataFiles/oreo1.dat>.

# Data and preparation

```{r data}
url <- "https://raw.githubusercontent.com/neilhatfield/STAT461/master/dataFiles/oreo1.dat"
oreo <- read.csv(url, check.names = TRUE)

# Put the groups in a meaningful order and use short labels in output.
oreo$Type <- factor(oreo$Type, levels = c("Regular", "Double Stuf"))
str(oreo)
head(oreo)
```

# Exploratory analysis

```{r summaries}
summary_by_type <- aggregate(Filling.Mass ~ Type, data = oreo,
                             FUN = function(x) c(n = length(x),
                                                 mean = mean(x),
                                                 sd = sd(x),
                                                 median = median(x),
                                                 min = min(x),
                                                 max = max(x)))
summary_by_type <- data.frame(Type = summary_by_type$Type,
                              do.call(rbind, list(summary_by_type$Filling.Mass)),
                              row.names = NULL)
knitr::kable(summary_by_type, digits = 3,
             caption = "Summary of filling mass (grams) by Oreo type.")

means <- tapply(oreo$Filling.Mass, oreo$Type, mean)
observed_ratio <- unname(means["Double Stuf"] / means["Regular"])
observed_ratio
```

```{r plot}
boxplot(Filling.Mass ~ Type, data = oreo,
        col = c("#8ECae6", "#FFB703"),
        ylab = "Filling mass (g)", xlab = "Oreo type",
        main = "Creme filling mass by Oreo type")
stripchart(Filling.Mass ~ Type, data = oreo, vertical = TRUE,
           method = "jitter", pch = 16, col = "gray25", add = TRUE)
```

The observed ratio of the two sample means is
**`r round(observed_ratio, 3)`**. Thus, the Double Stuf sample mean is about
**`r round(100 * (observed_ratio - 1), 1)`%** greater than the Regular sample
mean, rather than exactly 100% greater.

# Statistical analysis

To directly address “twice as much,” define the contrast

$$L = \mu_{\text{Double Stuf}} - 2\mu_{\text{Regular}}.$$

The hypotheses are

$$H_0: L = 0 \qquad \text{versus} \qquad H_A: L \ne 0.$$

A Welch two-sample *t* test is performed by comparing Double Stuf filling masses
to twice the Regular filling masses. Welch's test does not assume equal variances.

```{r hypothesis-test}
regular <- subset(oreo, Type == "Regular")$Filling.Mass
double_stuf <- subset(oreo, Type == "Double Stuf")$Filling.Mass

contrast_test <- t.test(double_stuf, 2 * regular,
                        alternative = "two.sided", var.equal = FALSE)
contrast_test

estimated_contrast <- mean(double_stuf) - 2 * mean(regular)
ci <- contrast_test$conf.int
```

The estimated contrast is **`r round(estimated_contrast, 3)` g**. The 95% confidence
interval for $$\mu_{\text{Double Stuf}} - 2\mu_{\text{Regular}}$$ is
(**`r round(ci[1], 3)`**, **`r round(ci[2], 3)`**) g, and the two-sided *p*-value is
**`r format.pval(contrast_test$p.value, digits = 3, eps = 0.001)`**.

# Conclusion

At the $$\alpha = 0.05$$ significance level, we
**`r if (contrast_test$p.value < 0.05) "reject" else "fail to reject"`** the null
hypothesis that Double Stuf Oreos contain exactly twice the *mean* creme filling
of Regular Oreos. In this sample, the estimated mean ratio is
**`r round(observed_ratio, 3)`**, so Double Stuf Oreos averaged somewhat less
than twice the creme filling mass of Regular Oreos.

This conclusion concerns the population represented by these sampled cookies and
their measured filling masses. It does not establish that every individual Double
Stuf Oreo has exactly twice the filling of every Regular Oreo.

# Reproducibility

```{r session-info}
sessionInfo()
```
