---
title: "Do Double Stuf Oreos Have Twice the Creme Filling of Regular Oreos?"
author: "Student D"
date: "9/4/2026"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")

# Packages
packages <- c("tidyverse", "knitr", "kableExtra", "car", "psych")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if (length(new_packages)) install.packages(new_packages)

library(tidyverse)
library(knitr)
library(kableExtra)
library(car)
library(psych)
```

# Introduction

Oreo cookies are marketed in a "Regular" form and a "Double Stuf" form, the
latter implying (via its name and marketing) that it contains twice the amount
of creme filling as the Regular version. This report uses measured filling
mass data (in grams) from a sample of Regular and Double Stuf Oreo cookies to
investigate the research question:

> **Do Double Stuf Oreos have twice the creme filling of Regular Oreos?**

We will explore the data, compare average filling mass between the two types,
and formally test whether the mean filling mass of Double Stuf cookies is
statistically consistent with being **exactly double** that of Regular
cookies.

# Data

The data come from a `.dat` file (comma-separated) hosted by Prof. Neil
Hatfield (Penn State, STAT 461) at:

`https://raw.githubusercontent.com/neilhatfield/STAT461/master/dataFiles/oreo1.dat`

Each row records the filling mass (grams) of a single cookie and its type
(`Regular` or `Double Stuf`).

```{r loadData}
dataURL <- "https://raw.githubusercontent.com/neilhatfield/STAT461/master/dataFiles/oreo1.dat"

oreoData <- read.csv(file = dataURL, header = TRUE, sep = ",",
                      stringsAsFactors = FALSE)

# Clean up column names for easier use
names(oreoData) <- c("fillingMass", "type")

oreoData <- oreoData %>%
  mutate(
    type = str_trim(type),
    type = factor(type, levels = c("Regular", "Double Stuf"))
  )

str(oreoData)
```

```{r previewData}
kable(head(oreoData, 10), caption = "First 10 rows of the Oreo filling data",
      col.names = c("Filling Mass (g)", "Type")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

# Exploratory Data Analysis

## Sample Sizes and Summary Statistics

```{r summaryStats}
summaryTable <- oreoData %>%
  group_by(type) %>%
  summarise(
    n = n(),
    Mean = mean(fillingMass),
    SD = sd(fillingMass),
    Median = median(fillingMass),
    Min = min(fillingMass),
    Max = max(fillingMass)
  )

kable(summaryTable, digits = 3,
      caption = "Summary statistics of filling mass (g) by Oreo type") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Visualizing the Distributions

```{r boxplot, fig.width=6, fig.height=4}
ggplot(oreoData, aes(x = type, y = fillingMass, fill = type)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red") +
  geom_jitter(width = 0.08, alpha = 0.4) +
  labs(
    title = "Filling Mass by Oreo Type",
    x = "Oreo Type",
    y = "Filling Mass (g)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r histogram, fig.width=7, fig.height=4}
ggplot(oreoData, aes(x = fillingMass, fill = type)) +
  geom_histogram(binwidth = 0.25, color = "black", alpha = 0.7, position = "identity") +
  facet_wrap(~type, ncol = 1, scales = "free_y") +
  labs(
    title = "Distribution of Filling Mass by Oreo Type",
    x = "Filling Mass (g)",
    y = "Count"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

*Note:* A few observations appear unusually extreme relative to their labeled
group (e.g., a very low value labeled "Double Stuf" and a very high value
labeled "Regular"). These may be data entry errors, mislabeled cookies, or
genuine measurement variability; they are retained here since we have no
independent way to verify or correct them, but they are worth flagging as a
limitation.

## Observed Ratio of Means

A simple, descriptive way to address the research question is to compute the
ratio of the mean filling masses:

```{r ratio}
means <- summaryTable$Mean
names(means) <- summaryTable$type

observedRatio <- means["Double Stuf"] / means["Regular"]
observedRatio
```

The Double Stuf cookies have, on average, about
**`r round(observedRatio, 2)`** times the filling mass of Regular cookies —
descriptively close to, but not exactly, double.

# Formal Analysis

To move beyond a purely descriptive comparison, we test whether the data are
consistent with the claim that Double Stuf filling mass is *exactly* double
the Regular filling mass, i.e., $\mu_{DS} = 2\mu_{R}$.

## Approach

We construct a new variable: for Regular cookies, we compare $2 \times$
filling mass against the Double Stuf filling mass directly, using a two-sample
$t$-test. Equivalently, we test:

$$H_0: \mu_{DS} - 2\mu_{R} = 0 \quad \text{vs.} \quad H_A: \mu_{DS} - 2\mu_{R} \neq 0$$

```{r checkAssumptions}
# Check variance homogeneity (using doubled Regular values for a fair scale comparison)
regularDoubled <- oreoData %>%
  filter(type == "Regular") %>%
  mutate(fillingMass = fillingMass * 2, type = "2x Regular")

dsData <- oreoData %>% filter(type == "Double Stuf") %>%
  mutate(type = as.character(type))

compData <- bind_rows(regularDoubled, dsData) %>%
  mutate(type = factor(type, levels = c("2x Regular", "Double Stuf")))

leveneTest(fillingMass ~ type, data = compData)
```

```{r ttest}
ttestResult <- t.test(
  fillingMass ~ type,
  data = compData,
  var.equal = TRUE,   # update to FALSE if Levene's test indicates unequal variances
  alternative = "two.sided"
)

ttestResult
```

```{r ttestTable}
ttestSummary <- tibble(
  `Mean (2x Regular)` = mean(regularDoubled$fillingMass),
  `Mean (Double Stuf)` = mean(dsData$fillingMass),
  `Difference` = ttestResult$estimate[2] - ttestResult$estimate[1],
  `t statistic` = ttestResult$statistic,
  `df` = ttestResult$parameter,
  `p-value` = ttestResult$p.value
)

kable(ttestSummary, digits = 4,
      caption = "Test comparing Double Stuf filling mass to 2x Regular filling mass") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Interpretation

- If the resulting **p-value is small** (e.g., below 0.05), we have evidence
  that the mean Double Stuf filling mass is *significantly different* from
  exactly twice the mean Regular filling mass — meaning the "double" claim
  does **not** hold exactly.
- If the p-value is **not small**, the data are consistent with the claim
  that Double Stuf filling is (on average) twice that of Regular.

```{r interpretationText}
pval <- ttestResult$p.value
diffEstimate <- ttestResult$estimate[2] - ttestResult$estimate[1]

conclusionText <- if (pval < 0.05) {
  paste0("The p-value (", format(pval, scientific = TRUE, digits = 3),
         ") is less than 0.05, so we reject the null hypothesis. ",
         "The mean Double Stuf filling mass is statistically significantly ",
         "different from exactly double the mean Regular filling mass ",
         "(estimated difference: ", round(diffEstimate, 3), " g).")
} else {
  paste0("The p-value (", round(pval, 4),
         ") is not less than 0.05, so we fail to reject the null hypothesis. ",
         "The data are consistent with Double Stuf filling being, on average, ",
         "double that of Regular filling.")
}

conclusionText
```

# Conclusion

Based on the sample data:

- The **observed ratio** of mean filling masses (Double Stuf / Regular) is
  approximately **`r round(observedRatio, 2)`**.
- The formal hypothesis test comparing Double Stuf filling mass to *exactly*
  twice the Regular filling mass yields `r conclusionText`

**Practical takeaway:** Even if the ratio is statistically different from
exactly 2, it may still be *practically* close to double, depending on the
magnitude of the difference and the context (e.g., manufacturing tolerances).
Readers should weigh both **statistical significance** and **practical
significance** when interpreting these results.

# Limitations

- A handful of observations appear mismatched with their labeled type (e.g.,
  an unusually small "Double Stuf" value and an unusually large "Regular"
  value), which may reflect data entry errors or true manufacturing
  variability. These were not removed, but a sensitivity analysis excluding
  them could be a useful extension.
- Sample sizes for each group are moderate (~30 each), which limits the power
  of the test to detect small departures from the exact "double" claim.
- Only one production batch/sample of cookies was measured; results may not
  generalize to all Oreo products or manufacturing runs.

# Appendix: Session Info

```{r sessionInfo}
sessionInfo()
```
