---
title: "Homework #1.1-Example A"
author: "A. Student"
date: "9/2/2022"
output: pdf_document
geometry: left=1in,right=1in,top=1in,bottom=1in
urlcolor: blue
header-includes:
  - \usepackage{subfig}
---

```{r setupFiles, include = FALSE}
# Setting Document Options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

# Add additional packages by name to the following list
packages <- c("tidyverse", "knitr", "kableExtra", "psych")
lapply(
  X = packages,
  FUN = library,
  character.only = TRUE
)

oreoData <- read.table(
  file = "https://raw.github.com/neilhatfield/STAT461/master/dataFiles/oreo1.dat",
  header = TRUE,
  sep = ","
)

```

<!-- This will be Beginning/Progressing Example -->

For this assignment, I was assigned the Oreo 1 data set.

Here are the values of statistics:

```{r statsR, echo=TRUE}
summary(
  oreoData[which(oreoData$Type == "Regular"), "Filling.Mass"],
  na.rm = TRUE
)
var(oreoData[which(oreoData$Type == "Regular"), "Filling.Mass"])
psych::skew(oreoData[which(oreoData$Type == "Regular"), "Filling.Mass"])
```

```{r statsD, echo=TRUE}
summary(
  oreoData[which(oreoData$Type == "Double Stuf"), "Filling.Mass"],
  na.rm = TRUE
)
var(oreoData[which(oreoData$Type == "Double Stuf"), "Filling.Mass"])
psych::skew(oreoData[which(oreoData$Type == "Double Stuf"), "Filling.Mass"])
```

The sample min is the smallest value, the sample max is the biggest value. Twenty-five percent of the cookies have 2.936 grams of filling (or 5.59). Half of the cookies have less than 3.04 grams (5.85 grams) of créme filling. The average of regular oreos is 3.178 grams; double stuf are 5.915 grams. Q3 for the oreos are 3.23 and 6.41. The variances are the square of the standard deviation, so they are the squared typical deviation from the average. Skewness tells us that the regular oreos are left skewed while double stuff are right skewed as shown in the histograms.

```{r hists}
hist(
  x = oreoData[which(oreoData$Type == "Regular"), "Filling.Mass"]
)

hist(
  x = oreoData[which(oreoData$Type == "Double Stuf"), "Filling.Mass"]
)

```

The question of whether double stuf oreos are actually double the stuff, I guess, is a two sample location problem. However, I don't really know.

The null hypothesis would be 2D = R; while the alternative hypothesis would be 2D not equal R. This doesn't seem right, I'm used to things like mu = 0 and mu1 - mu2 = 0. I suppose I could do 2D - R = 0. I'll do a Two Sample T Test.

```{r tTest, echo=TRUE}
t.test(
  x = oreoData[which(oreoData$Type == "Double Stuf"), "Filling.Mass"],
  y = oreoData[which(oreoData$Type == "Regular"), "Filling.Mass"],
  conf.level = 0.97
)
```

My t value is 15.661; degrees of freedom is 57.566. The p-value is really small (< 2.2e-16). This would mean that there is essentially no chance that the null hypothesis is true. The 97% confidence interval is (2.348, 3.125); we're 97% confident that the true difference between double sutf and regular oreos is between 2.348 and 3.125 grams. The mean of regular oreos is 3.178, which is not in this interval; so clearly double stuf can't have double the filling; they have less.

\newpage

# Code Appendix

```{r codeAppendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```
