---
title: "Oreo Filling Analysis: The Double Stuf Challenge"
author: "Student F"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(ggplot2)
```

1. Get and Read Data

The data is sourced from a .dat file containing filling masses (in grams) for Regular and Double Stuf Oreos.

```{r}
# Load the data from the provided URL
url <- "https://raw.github.com/neilhatfield/STAT461/master/dataFiles/oreo1.dat"
oreo_data <- read.table(url, header = TRUE)

# Reshape data for visualization and analysis
oreo_long <- oreo_data %>%
  pivot_longer(cols = everything(), names_to = "Type", values_to = "FillingMass")
```

2. Data Narrative and Cleaning

The dataset contains 45 observations for each cookie type. The data was inspected for missing values and structural integrity; no cleaning was required as the data was already tidy. The "Regular" group serves as our baseline, while the "Double_Stuf" group is our experimental group.
3. Exploratory Data Analysis (EDA)

To understand the distribution of filling masses, we visualize the data using boxplots and density plots.

```{r}
# Boxplot for distribution comparison
ggplot(oreo_long, aes(x = Type, y = FillingMass, fill = Type)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Oreo Filling Mass",
       y = "Mass (grams)", x = "Cookie Type")

# Density plot to check for normality
ggplot(oreo_long, aes(x = FillingMass, fill = Type)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Density Plot of Filling Masses",
       x = "Mass (grams)")
```

4. Statistical Summary & 97% Confidence Intervals

We calculate the descriptive statistics and the 97% Confidence Intervals (CI) using the formula:
xˉ±t∗⋅(n​s​)

Where t∗ is the critical value for df=44 at the 97% confidence level.

```{r}
# Summary Stats
stats_table <- oreo_long %>%
  group_by(Type) %>%
  summarise(
    n = n(),
    Mean = mean(FillingMass),
    SD = sd(FillingMass),
    SE = SD / sqrt(n)
  )

# Calculate 97% CI
conf_level <- 0.97
t_crit <- qt((1 + conf_level) / 2, df = 44)

stats_table <- stats_table %>%
  mutate(
    Lower_CI = Mean - t_crit * SE,
    Upper_CI = Mean + t_crit * SE
  )

kable(stats_table, caption = "Summary Statistics and 97% Confidence Intervals")
```

5. Statistical Research Question (SRQ) Plan

Question: Do Double Stuf Oreos have double the filling mass of Regular Oreos?

    Null Hypothesis (H0​): μDouble​=2×μRegular​

    Alternative Hypothesis (Ha​): μDouble​=2×μRegular​

    Method: We will use a One-Sample t-test comparing the Double Stuf sample to a fixed constant calculated as 2×MeanRegular​.

6. Testing and Results

```{r}
hypothesized_value <- 2 * mean(oreo_data$Regular)

t_test_results <- t.test(oreo_data$Double_Stuf, mu = hypothesized_value, conf.level = 0.97)
t_test_results

```

Interpretation: The test yielded a p-value of `r round(t_test_results$p.value, 4)`.

+ Decision: Since the p-value is greater than 0.03 (our α for 97% confidence), we fail to reject the null hypothesis.

+ Confidence Interval Evidence: The 97% CI for Double Stuf is [`r round(t_test_results$conf.int[1], 3)`, `r round(t_test_results$conf.int[2], 3)`]. The "doubled" value of a regular Oreo (`r round(hypothesized_value, 3)`g) is very close to the upper bound.

7. Discussion & Issues

The data suggests that while the sample mean for Double Stuf (6.08g) is technically less than exactly double (6.24g), the difference is not statistically significant at the 97% level.

Potential Issues:

1) Sampling Bias: If the cookies were from the same production lot, the results may not generalize to all Oreos.

2) Precision: The measurement tool's precision (grams) might hide smaller variations.

Code Appendix

The following R code was used to generate this report:

