---
title: 'HW #1.1: Exploring Oreo Data (Baseline)'
author: "Available: Aug. 24, 2022; Due: Friday, Sept. 2, 2022"
date: '`r paste("Last Updated:", Sys.Date())`'
output: pdf_document
geometry: left=0.5in,right=0.5in,top=0.5in,bottom=0.5in
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```

There are __TWO (2)__ parts to this assignment: value checking via MyOpenMath and a Written Report.

For this homework assignment you will need to use the Oreo Data Set that has been assigned to you from inside the HW 1.1: Oreo Comparison-Value Check (MyOpenMath Assignment linked in Canvas). The first question will have which data file you've been assigned (Oreo 1, Oreo 2, or Oreo 3) and will display the link to the data.

# Background

Food product marketing often invokes strong feelings from consumers. For example, Subway was sued by consumers over their "foot-long" sub sandwiches not being 1 foot/12 inches long. In a similar vein, we might ask whether the marketing around one version of a product being "double" another version is accurate. Thus was born an investigation.

In 2017, I worked with an Introductory Statistics class to test out whether Double Stuf Oreos have twice the amount of stuf as Regular Oreos. ("Stuf" is the official term for the créme filling.) We designed a stochastic process that allowed us to 1) select a grocery store in Tempe, AZ, via a lottery, 2) conduct a stratified random sampling of Regular and Double Stuf Oreo packages (9 packages each), and 3) weigh the amount of créme filling (in grams) for all cookies in each package. we separaated the packages into 9 separate pairings (one for each team) of Double and Regular for data collection and analysis.

In this assignment, you will explore the same question as these students did (i.e., Do Double Stuf Oreos have double the créme filling mass as Regular Oreos?). Each student will be assigned one of three selected data sets. I will use another data set as a demo in class. 

*Caution:* don't assume that the data is error free.

# Task Summary

For this assignment, you'll need to complete the following tasks. [MOM] indicates that there is something you need to do in the MOM portion of the assignment; [Report] indicates that there is something you need to in your final report; [MOM/Report] indicates that you need to do something in both the MOM and Report portions of the assignment.

1. Get and Read in the Appropriate Data using the software of your choice [MOM]
2. Explore the Data and Write a Data Narrative 
    a. Clean your data
    b. Make appropriate data visualizations [Report]
    c. Find and interpret the values of various statistics [MOM/Report]
    d. Write your data narrative [Report]
3. Develop and Carry Out a Plan to Answer the SRQ
    a. State the Hypotheses/Models you will test [MOM/Report]
    b. Explain how you will test and *why* this test is appropriate [Report]
    c. Explain the results of the test [MOM/Report]
    d. State your decision [MOM/Report]
    e. Discuss any issues [Report]
4. Code Appendix--won't be graded, only examined [Report]

What follows are some notes (and hints) for approaching each of the tasks.

\newpage

# Task 1-Read in the Data.

Your first task to get the data. Again, you'll need to check out the first question of the MOM component to see which data set you've been assigned. Be sure that you click the appropriate choice in the question to ensure that the rest of the questions are graded appropriately.

Once you have the data (you've downloaded the data or you've copied the URL to the file), you'll need to open it up in your chosen software. 

Your first task will be to read the data into your chosen software program. If you are using something other than R, I might suggest that you copy/paste the appropriate URL into your browser and download the data file. Then open that in your selected program. Several software packages will import the data for you from the URL.

## Hint 1--Use the URL to Read in Data.

If you are using __JMP__, select Internet Open... from the File menu and then paste the URL into the box and then click OK.

If you are using __R__, then you can read the data with the following:
```{r readData, echo=TRUE}
demoOreo <- read.table(
  file = "https://raw.github.com/neilhatfield/STAT461/master/dataFiles/classDemoOreo.dat", 
  header = TRUE,
  sep = ","
)
```

Note: you'll need to replace the example URL with yours.

# Task 2-Explore your data and write a data narrative.

Explore the data using your selected software package. Your answer here should come in form of a data story, complete with data visualizations, interpretations of various statistics, and plenty of discussion of both. 

## Hint 2

Be sure that you have checked your data for any cleaning that should be done. Look at data visualizations. Are there any cookies that seem to not make sense with the rest of their type? Anything looked switched? Be sure you describe any such issues and the remedy you implemented in your narrative.

## Hint 3
When making data visualizations, be sure that you use an appropriate title and label the axes. Some useful commands to add to a `ggplot` call:

+ `ggtitle("your title here")`
+ `xlab("label for horizontal axis")`
+ `ylab("label for vertical axis")`
+ `theme_*()` (invokes a pre-built style; several options, plus you can build your own)

## Hint 4
Control the size and position of your plots by adding options to code chunk header. 

That is, `{r chunkName, option1=, option2=, etc.}`

+ `fig.align='left/center/right'`
+ `fig.height=7` and `fig.width=7` (in inches)

## Hint 5
You can reference plots in your narrative in a couple of easy steps. 

### R Markdown
1. Build your code chunk to make a plot and give the chunk a name.
2. In the chunk options, you'll need to define a caption: `fig.caption="caption text goes here"` (If you do this, you can omit the graph title.)
2. In your text type `\ref{fig:chunkName}` to call the reference to your plot.

For example, if I want to talk about the scatterplot (Figure \ref{fig:scatterPlot}), I can have the system keep track of the numbering for me.

```{r scatterPlot, fig.align="center", fig.width=3, fig.height=3, fig.cap="Scaterplot of Filling Mass (g) by Type of Oreo"}
ggplot(demoOreo, aes(x=Type, y=Filling.Mass)) +
  geom_point() +
  theme_bw() +
#  ggtitle("Filling Mass (g) by Type") +
  xlab("Type of Oreo") +
  ylab("Filling Mass (g)")
```

### Microsoft Word-Current Version
1. Insert your graphic into Word.
2. Click/Select your graphic, and then from the Insert Menu select Caption...
3. Fill in the appropriate information and press OK.
4. In a sentence where you want to add a reference, select Cross-reference... from the Insert Menu
5. Make the appropriate selections in the drop down menu and selection field. Press Insert when finished.

*Note: Word will update the numbering when you call up the Print menu (even if you don't print). You'll need to do this if you change the order of your figures, add/remove some, etc.

## Requested statistics.

You must include the following statistics in your narrative. You may include additional statistics as you wish. However, __you must provide an interpretation of one value from each statistic__. Please include the appropriate units where applicable. You may place them in a table BUT you must also incorporate them into your data narrative.

| statistic | Regular Oreos | Double Stuf |
|:----------|:-------------:|:-----------:|
|*Sample Max*|    |    |
|*Third Quartile*|    |    |
|*Sample Median*|    |    |
|*First Quartile*|    |    |
|*Sample Min*|    |    |
|*Sample Arithmetic Mean*|    |    |
|*Sample Arithmetic Variance*|    |    |
|*Sample Skewness*|    |    |

\newpage

# Task 3-Develop and Carry Out a Plan to Answer the SRQ

In this task, you'll need to explain how you're going to approach answering the SRQ using the tools of hypothesis testing. Then, you'll need to conduct a test and report all appropriate values. Finally, you'll need to discuss these results.

You will need to report any values of a test statistic (as well as degrees of freedom, if applicable), your *p*-value, and a __97%__ confidence interval. You must interpret both the *p*-value and the confidence interval in your report.

## Hint 6

Challenge yourself here. Think through what would be good methods to answer the question. 

## Hint 7

You know what they say about assumptions...CHECK THEM! In your narrative, discuss the assumptions, how you checked each one, and what you learned by checking them.

## Hint 8

Type set your mathematical statements. In Microsoft Word use the Equation Editor. In R Markdown, use LaTeX expressions.

# Task 4-Code Appendix

This will help us see if there are any issues with the code you ran (i.e., why your values might be incorrect). Your choice of software package will dictate what you put here.

## GUI Bases Package--SPSS, JMP, Minitab, etc.
Please include a statement of which package you used, including version number (check out the "About" menu option for information). You may also copy your log/journal to the appendix.

Write a brief description of what parts of the package you used, but do not give a click-by-click accounting.

## Command Line Package--R, SAS, etc.

Include the actual code you used, being sure to comment and organize your code.

### Hint 9

Use the [Homework Template RMD file](https://www.dropbox.com/s/cd70l4szvqu9k1x/hwtemplate.rmd?dl=0) (Also on [GitHub](https://github.com/neilhatfield/STAT461/blob/master/hwTemplate.Rmd).) to automatically create a code appendix for you.