---
title: "Parametric Shortcut for One-way ANOVA"
author: "Neil J. Hatfield"
date: "2/24/2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to conduct the parametric shortcut known as the
  One-way ANOVA *F* Test. This will include assessing whether
  the situation is amenable to ANOVA as well as assessing the 
  assumptions for the parametric shortcut.
---
```{r setup0, include=FALSE}
library(learnr)

learnr::tutorial_options(exercise.timelimit = 60)

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  fig.align = "center"
)

packages <- c("tidyverse", "hasseDiagram", "knitr", 
              "kableExtra", "car", "psych",
              "parameters")
lapply(packages, library, character.only = TRUE)

# Tell Knitr to use empty space instead of NA in printed tables
options(knitr.kable.NA = "")


```

In this tutorial, we are going to explore using R to fit a One-way ANOVA model to two data sets using the parametric shortcut known as the One-way ANOVA *F* Test.

# Prepare R

The first thing that we need to do is to make sure that R is ready for us to do our work. There are two aspects to this: loading packages and putting in our constraint.

## Loading Packages

For this tutorial we are going to need the following packages:

+ For Analysis: `tidyverse`, `hasseDiagram`, `car`, `psych`, and `parameters`
+ For Professionalism: `knitr` and `kableExtra`

Use the code block below to load these packages into your session.

```{r loadPackages, exercise=TRUE}

```

```{r loadPackages-solution, exercise.reveal_solution = TRUE}
packages <- c("tidyverse", "hasseDiagram", "knitr", 
              "kableExtra", "car", "psych",
              "parameters")
lapply(packages, library, character.only = TRUE)

# Note you could also use individual library calls for each package

```

## Set Our Constraints

Recall that in order to ensure that we have estimable functions, we must set a side condition or constraint on our treatments: 
\[\sum_i^k\alpha_i=0\]
This is not what R does by default, but we can tell R to adopt this constraint with the following code:

```{r setConstraint, echo=TRUE}
options(contrasts = c("contr.sum", "contr.poly"))

```

Now that we have both of these, we can turn our attention to the next step.

# Load Data

Our first step will be to load data into our R session. We are going to use two data sets: the honey example from class and Example 3.2 Resin Lifetimes from the Oehlert textbook. For the honey data, we will create the data frame manually; for resin lifetimes, we'll import the data.

```{r getData, exercise=TRUE}
## DEMO CODE

# Honey Data
honey <- data.frame(
  Amount = c(150, 50, 100, 85, 90, 95, 130, 50, 80),
  Varietal = rep(c("Clover", "Orange Blossom", "Alfalfa"), each = 3)
)
## Set Varietal to factor
honey$Varietal <- as.factor(honey$Varietal)

# Resin Lifetimes Data
resin <- oreoData <- read.table(
  file = "https://raw.github.com/neilhatfield/STAT461/master/dataFiles/resinLifetimes.dat",
  header = TRUE,
  sep = ""
)
## Set temp to factor
resin$temp <- as.factor(resin$temp)

```