---
title: "Mealybugs on Cycads Solution"
author: "Neil J. Hatfield"
date: "March 29, 2021"
output: pdf_document
geometry: left=1in,right=1in,top=1in,bottom=1in
urlcolor: blue
header-includes:
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{xcolor}
---

```{r setupFiles, include = FALSE}
# Setting Document Options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

# Add additional packages by name to the following list
packages <- c("tidyverse", "knitr", "kableExtra",
              "hasseDiagram", "car", "psych", "parameters",
              "DescTools") 
lapply(
  X = packages,
  FUN = library,
  character.only = TRUE
)

# Set constraint
options("contrasts" = c("contr.sum", "contr.poly"))

options(knitr.kable.NA = "")
source("https://raw.github.com/neilhatfield/STAT461/master/rScripts/ANOVATools.R")

```

Use this activity sheet to guide you through practicing designing a Randomized Complete Block Design (RCBD).

# The Context

Modern zoos try to reproduce natural habitats in their exhibits as much as possible. They therefore use appropriate plants, but these plants can be infested with inappropriate insects. Zoos need to take great care with pesticides, because the variety of species in a zoo makes it more likely that sensitive species is present.

Cycads (plants that look vaguely like palms) can be infested with mealybug and the zoo wishes to test three treatments: water (a null treatment/placebo), horticultural oil (a standard no-mammalian-toxicity pesticide), and fungal spores in water (*Beauveria bassiana*, a fungus that grows exclusively on insects). Five infested cycads are removed to a testing area. Three branches are randomly chosen on each cycad and two 3cm by 3cm patches are marked on each branch; the number of mealybugs in these patches is noted. The three branches on each cycad are randomly assigned to the three treatments. After three days, the patches are counted again, and the response is the change in the number of mealybugs (before -- after).

## What are the Statistical Research Questions?

[Develop at least two SRQs for this study]

# Study Design

With our SRQs in mind, let us design a study which will allow us to answer the SRQ.

## Key Elements

+ Response: [fill in]
    - Unit of Measure: [fill in]
+ Measurement Unit: [fill in]
+ Factor: [fill in, fixed or random effect?]
+ Block: [fill in]
+ Experimental Design:
    1) Experimental Units: [fill in]
    2) Treatments: [fill in]
    3) Treatment Selection Method: [fill in]
    4) Treatment Assignment Method: [fill in]
+ What kind of design? Randomized Complete Block Design

## Study Design Description

[Transform the Key Elements list into a descriptive narrative of how you will conduct the study]

# Exploratory Data Analysis

```{r loadData}
# Data
mealybugs <- data.frame(
  change = c(
    -9, 18, 3, 9, 5,
    -6, 5, 9, 0, 13,
    -4, 29, 4, -2, 11,
    7, 10, -1, 6, -1,
    4, 29, 14, 14, 7,
    11, 36, 16, 18, 15
  ),
  plant = as.factor(
    rep(1:5, times = 6)
  ),
  treatment = as.factor(
    rep(
      x = c("Water", "Spores", "Oil"),
      each = 10
    )
  )
)

```

[Your Data Explorations Go Here]

```{r exploration1}
# You'll have several code blocks here as you engage in EDA.
# You should create several data visualizations as well as
# tables of summary statistics.
```

# Statistical Inference Methods

```{r hasseDiagram, fig.cap="Hasse Diagram for Mealbug Treatment Study", fig.height=2}
# Create your Hasse Diagram here
modelLabels <- c("1 Kill Bugs 1", "5 Plant 4", "3 Pest Treatment 2", "30 (Patches) 23")
modelMatrix <- matrix(
  data = c(FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE),
  nrow = 4,
  ncol = 4,
  byrow = FALSE
)
hasseDiagram::hasse(
 data = modelMatrix,
 labels = modelLabels
)

```

[Explain how our study does or does not meet the base requirements of One-way ANOVA.]

## Hypotheses

[State your hypotheses here.]

## Method

To decide between the hypotheses and answer our research question, we will use [insert method choice here]. To account for multiple comparison problems, we will conceptualize the following testing family(-ies):

+ [describe testing family]

Further, we will control the [name] Type I Error Rate, with an overall Type I risk of [value]. We will achieve this by using the [name] method.

```{r inference1}
bugModel <- aov(
  formula = change ~ treatment + plant,
  data = mealybugs
)

```

## Assumptions

[List and discuss the assumptions of your selected method to get to a sampling distribution]

```{r assumptions1}
car::qqPlot(
  x = residuals(bugModel), 
  distribution = "norm",
  envelope = 0.90,
  id = FALSE,
  pch = 20,
  ylab = "Residuals (bugs)"
)

```

```{r assumptions2}
ggplot(
  data = data.frame(
    residuals = residuals(bugModel),
    fitted = fitted.values(bugModel)
  ),
  mapping = aes(x = fitted, y = residuals)
) +
  geom_point(size = 2) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_smooth(
    formula = y ~ x,
    method = stats::loess,
    method.args = list(degree = 1),
    se = FALSE,
    size = 0.5
  ) +
  theme_bw() +
  xlab("Fitted values") +
  ylab("Residuals")

```

```{r interactionPlot}
ggplot(
  data = mealybugs,
  mapping = aes(
    x = treatment,
    y = change,
    group = plant,
    color = plant
  )
) +
  geom_point() +
  stat_summary(fun = mean, geom = "line") +
  theme_bw() +
  xlab("Treatment") +
  ylab("Change in Bug Count") +
  labs(color = "Plant") +
  theme(
    legend.position = "right"
  )

```


# Results

```{r results1}
parameters::model_parameters(
    model = bugModel,
    omega_squared = "partial",
    eta_squared = "partial",
    epsilon_squared = "partial"
) %>%
  knitr::kable(
    digits = 4,
  col.names = c("Source", "SS", "df", "MS", "F", "p-value",
                "Partial Omega Sq.", "Partial Eta Sq.", "Partial Epsilon Sq."),
  caption = "ANOVA Table for Mealybug Pesticide Study",
  align = c('l',rep('c',8)),
  booktab = TRUE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "condensed"),
    font_size = 10,
    latex_options = c("scale_down", "HOLD_position")
  )

```


```{r results2}
bugPH <- TukeyHSD(
  x = bugModel,
  which = "treatment",
  conf.level = 0.97
)

knitr::kable(
  bugPH$treatment,
  digits = 4,
  caption = "Post Hoc Tukey HSD Comparisons",
  col.names = c("Difference", "Lower Bound",
                "Upper Bound", "Adj. p-Value"),
  align = 'cccc',
  booktabs = TRUE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "boardered"),
    font_size = 12,
    latex_options = "HOLD_position"
  )

```

```{r effectSizes}
anova.PostHoc(
  aov.obj = bugModel,
  response = "change",
  mainEffect = "treatment"
) %>%
  knitr::kable(
    digits = 3,
    caption = "Post Hoc Comparison Effect Sizes",
    col.names = c("Pairwise Comparison","Cohen's d", "Hedge's g",
                  "Prob. Superiority"),
    align = 'lccc',
    booktabs = TRUE
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "boardered"),
    font_size = 12,
    latex_options = "HOLD_position"
  )

```

[Write Up your results]

# Discussion

[Discuss the implications of the results section in the context of the study. What will you tell the client?]

\newpage

# Code Appendix

```{r codeAppendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```
